---
title: "Reproducible Research Project 1"
author: "Arga"
date: "01 November 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```
# Reproducible Research Project 1

This is an R Markdown file to report the answers and findings of the 1st assignment in Reproducible Research course by Johns Hopkins University. Each question is listed as a sub-heading in this document.

## Loading and preprocessing the data

In this report, we use ```ggplot2``` package to plot all the graphs.

```{r Load packages, results='hide'}
library(ggplot2)
library(dplyr)
```

To improve reproducibility, the environment parameters of analysis is written down here.

```{r Print environment parameters}
sessionInfo()
```


To start our analysis, we download the data from the web and put it in R data frame ```activity```. Download time is noted in ```downloadTime```
```{r Load the data}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip","Factivity.zip")
downloadTime <- Sys.time() ##Record the time of download
unzip("Factivity.zip")
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
activity$date <- as.Date(activity$date)
names(activity) <- c("Steps","Date","Interval")
str(activity)
```


## What is mean total number of steps taken per day?
As part of the instructions, missing value in the dataset is just left as it is. The following histogram shows the total steps taken in any given day. Days with total 0 steps are those that consist of missing values. 
Ignore the warning that comes after the code.

```{r Calculate steps and plot}
totalStepsEachDate <- aggregate(Steps~Date,data = activity,FUN = sum,na.rm=TRUE)
ggplot(data = totalStepsEachDate, aes(Steps)) +
        geom_histogram(bins = 35) +
        ylab("Frequency (days)") +
        ggtitle("Total Steps on Each Day")
stepMean<-mean(totalStepsEachDate$Steps)
stepMedian<-median(totalStepsEachDate$Steps)
```

The mean of total steps on each day is `r sprintf("%.2f",stepMean)` steps.
The median of total steps on each day is `r stepMedian` steps.

## What is the average daily activity pattern?
Now, we look at the activity pattern that emerges by averaging the values in each interval.

```{r Plotting time series of daily pattern}
averageDailySteps<-aggregate(Steps~Interval, data = activity, FUN = mean, na.rm = TRUE)
ggplot(averageDailySteps, aes(Interval, Steps)) +
      geom_line() +
      ggtitle("Average Daily Activity Pattern")

maxInterval<-averageDailySteps[which.max(averageDailySteps$Steps),]
```

The 5-minute interval containing the highest number of steps on average is the `r maxInterval[[1]]`th which has `r sprintf("%.f",maxInterval[[2]])` steps. 


## Imputing missing values

In our dataset, there are missing values in the `Steps` column. This section provides the method we use to impute the NAs.
```{r Take care of missing values}
table(complete.cases(activity))[[1]]
imputedSteps<-ifelse(is.na(activity$Steps),averageDailySteps[match(x = activity$Interval,table = averageDailySteps$Interval),2],activity$Steps)
activity2 <- data.frame(Steps=imputedSteps,Date=activity$Date,Interval=activity$Interval)
```

To be exact, there are `r table(complete.cases(activity))[[1]]` NAs in the dataset. We substitute the missing `Steps` value in the `activity` data frame with the average `Steps` value from the average daily pattern (`averageDailySteps`), and the imputed data frame is stored in `activity2`.

After imputation, the following plot of total steps on each day is generated from the `activity2` data frame.

```{r Examine the new dataset}
totalStepsEachDate2 <- aggregate(Steps~Date,data = activity2,FUN = sum,na.rm=TRUE)
ggplot(data = totalStepsEachDate2, aes(Steps)) +
        geom_histogram(bins = 35) +
        ylab("Frequency (days)") +
        ggtitle("Total Steps on Each Day")
stepMean2<-mean(totalStepsEachDate2$Steps)
stepMedian2<-median(totalStepsEachDate2$Steps)
```

Notice that there is no day with total 0 step (the minimum is `r min(totalStepsEachDate2$Steps)`).

The mean of total steps on each day is `r sprintf("%.2f",stepMean2)` steps.
The median of total steps on each day is `r sprintf("%.2f",stepMedian2)` steps. Our method of imputation does not change the aforementioned statistics by much. However, the estimate of total daily number of steps on days with previously missing values are all the same i.e. `r sprintf("%.2f",sum(averageDailySteps$Steps))` steps. This results in more days being counted in the bin with the average value in it.


## Are there differences in activity patterns between weekdays and weekends?

In the last part of our analysis, we like to see if there is any difference in activity on weekday and weekend. The following plot shows time-series graphs of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r Separate and analyse weekday and weekend activities}
dayType<-as.factor(ifelse(weekdays(activity$Date,abbreviate = TRUE)%in%c("Sat","Sun"),"Weekend","Weekday"))
activity2$Day.Type <- dayType
activity2 <- activity2 %>% group_by(Day.Type) %>% mutate(Mean=mean(Steps))

ggplot(data = activity2, aes(y = Steps, x = Interval)) +
    stat_summary(fun.y = mean, geom ="line") +
    facet_wrap(~Day.Type,nrow = 2) +
    geom_hline(aes(yintercept=Mean,group = Day.Type), colour='blue')
```

As you can see from the graphs, the highest number of steps happen during weekdays. The blue lines indicating the means tell us that weekend has higher average number of steps than weekdays.

